---
author: "Ryan Scehovic"
title: "DS 202 - Lab 4"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(readxl)
library(tidyverse)
```

Part 1 - Number 1
```{r}
offense <- read_excel('../ds202_lab4/cyclonesFootball2020.xlsx', sheet='Offensive')
defense <- read_excel('../ds202_lab4/cyclonesFootball2020.xlsx', sheet='Defensive')
biography <- read_excel('../ds202_lab4/cyclonesFootball2020.xlsx', sheet='Biography')
head(offense)
head(defense)
head(biography)

offense <- rename(offense, 'Opponent'='Opponent_Opponent')
defense <- rename(defense, 'Opponent'='Opponent_Opponent')

offense$Opponent <- as.factor(offense$Opponent)
defense$Opponent <- as.factor(defense$Opponent)

offense$Name <- as.factor(offense$Name)
defense$Name <- as.factor(defense$Name)

biography$Name <- as.factor(biography$Name)
```

Part 1 - Number 2
```{r}
offense <- offense %>% mutate(across(Receiving_REC:Passing_INT, as.numeric))
defense <- defense %>% mutate(across(Tackles_Solo:Pass_PB, as.numeric))
str(offense)
str(defense)

defClean <- defense
offClean <- offense
```

Part 1 - Number 3
```{r}
biography <- separate(biography, Height, c('Feet', 'Inches'), sep='-')
biography <- biography %>% mutate(across(Feet:Weight, as.numeric))
bioClean <- biography
str(biography)
```

Part 2 - Number 1
```{r}
offClean1 <- offClean %>% pivot_longer(Receiving_REC:Passing_INT,names_to = 'stat',values_to = 'value' )
offClean2 <- drop_na(offClean1)
```

Part 2 - Number 2
```{r}
offClean2$stat <- as.factor(offClean2$stat)
ggplot(offClean2, aes(x=value))+geom_histogram(bins = 15)+facet_wrap(~ stat, scales = 'free')+ggtitle("Team Stats")
```

Looking at the graphs above, Passing_INT and passing_TD both have a mean of 0, both are skewed to the right, passing_INT is more skewed to the right though. I also see a lot of players with 0 passing yards and only a few that have have than a 100. This is likely due to there only being a few starting wide recivers and some that rotate in and out, so there isn't enough playing time to have a lot of wide receivers get a lot of time. They only play the best ones. 

Part 2 - Number 3
```{r}
offCleanW<- offClean2 %>% filter(stat == "Receiving_YDS" & ( Opponent == "Oklahoma" | Opponent == "Oregon")) %>% pivot_wider(names_from='Opponent',values_from="value", values_fn=mean) 
 
head(offCleanW)
ggplot(offCleanW, aes(x=Oklahoma, y= Oregon))+geom_point()+xlim(0,100)+ylim(0,100)+ggtitle("Player Receiving Yards (Oklahoma vs. Oregon)")
```

Defense was better again Oklahoma because players had higher receiving yards when playing them and less against Oregon.

Part 2 - Number 4
```{r}
bioClean <- separate(bioClean, Hometown, c('City', 'State'), sep=', ')
head(bioClean[7:8], 5)
```

Part 2 - Number 5
```{r}
n5 <- table(bioClean$State)
n5 <- data.frame(rbind(n5))
n5
```

Part 3- Number 1
```{r}
homeCount <- n5 %>% pivot_longer(Ariz.:Wis., names_to='State', values_to='Count') %>% arrange(desc(Count))
homeCount
ggplot(bioClean, aes(x=`State`)) + geom_bar() + theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjus=.5))+ggtitle("Players per State")
```

Part 3- Number 2
```{r}
Brock <- offClean2 %>% filter(Name == "Purdy, Brock")
str(Brock)
ggplot(Brock, aes(x=value))+geom_histogram(bins = 15)+facet_wrap(~ stat, scales = 'free')+ggtitle("Brock Purdy Statistics")
```

The graph above shows how Brock Purdy influences and contributes to stats like interceptions, passing/rushing touchdowns, passing yards, rushing attempts/yards. 

```{r}
ggplot(offClean2, aes(x=value, fill=Name=="Purdy, Brock"))+geom_histogram(bins = 15)+facet_wrap(~ stat, scales = 'free')+ggtitle("Team Stats with Brock Purdy")
```

The graph aboves shows that Brock Purdy has a big influence on how the overall ofense does. A bad game from him could mean a bad game for the offense. This is seen in stats like passing yards, touchdowns, and rushing yards, where Purdy is important. 

Part 3- Number 3
```{r}
offense2019 <- read_excel('../ds202_lab4/cyclonesFootball2019.xlsx', sheet='Offensive')
offense2019 <- rename(offense2019, 'Opponent'='Opponent_Opponent')
offense2019$Opponent <- as.factor(offense2019$Opponent)
offense2019$Name <- as.factor(offense2019$Name)
offense2019 <- offense2019 %>% mutate(across(Rushing_ATT:Passing_INT, as.numeric))
offClean2019 <- offense2019
offenseClean2019_219 <- offClean2019 %>% pivot_longer(Rushing_ATT:Passing_INT, names_to = 'stat',values_to = 'value' )
offClean2019_2 <- drop_na(offenseClean2019_219)
offenseAvg <- offClean2 %>% group_by(Name, stat)%>% summarise_at(vars(value), list(avg2020 = mean))
offenseAvg2019 <- offClean2019_2 %>% group_by(Name, stat)%>% summarise_at(vars(value), list(avg2019 = mean))  
offenseBothYears <- offenseAvg %>% left_join(offenseAvg2019, by=c('Name', 'stat')) %>%
  drop_na() %>% mutate(difference = avg2020 - avg2019) %>% mutate(improved = difference>0)
head(offenseBothYears, 5)

ggplot(offenseBothYears, aes(x=Name, y=difference))+ geom_bar(stat = "identity")+ 
  facet_wrap(~stat, scales = 'free')+ ggtitle("Stat Averages (2019 - 2020)") + 
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjus=0.5, size = rel(0.7)))
```

Breece Hall looks like he imrpoved the most from 2019 to 2020 because you can see his difference in stats from the 2019 season to the 2020 season is the biggest of all players. You can see him compared to other players and no one else is close in Rushing_ATT/TD/YARDS. 